Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-07 paper/main.tex
shortened section 4

In the rule {\sc L:Skip}, we do not have to account for any resource
consumption.  As a result, the precondition $Q$ can be any (potential)
function and we only have to make sure that we do not end up with more
potential.  Since the execution of \code{skip} leaves the program
state unchanged, we can simply use the precondition as postcondition.
The potential functions $B$ for the \code{break} and $R$ for the
\code{return} part of the postcondition are not reachable and can
therefore be arbitrary.

@@

The rule {\sc L:Update} is the standard assignment rule of Hoare
logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
the precondition we ensure that $Q$ evaluates to the same number as in
the postcondition.  We also require that have we the constant
potential $M_u + M_e(e)$ available in the precondition to pay for the
cost of the evaluation of $e$ and the update.
\iffull{
}{}
The rule {\sc L:Seq} rule is crucial to understand how the
quantitative Hoare logic works.  To account for early exits of
statements, we must ensure in the \code{break} part $B$ of $S_1$'s
judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
same is true for the return part $R$ of the judgements for $S_1$ and
$S_2$.  The interaction between the actual pre- and postconditions is
analogous to standard Hoare logic.  In the postcondition of $S_1$ we
account for the cost $M_s$ of the execution of the sequence.

In the rule {\sc L:Loop}, the \code{break} part of the
loop body $S$ becomes the postcondition of the loop statement. We use
an arbitrary $B$ as the \code{break} part of the judgement for
$\code{loop}\, S$ since its operational semantics ensures that it can
only terminate with a \code{skip} or a \code{return}.  The
precondition $I$ of the loop is the loop invariant.  That is why we
require to have potential $I$ available in the precondition of the
loop body $S$.  In the postcondition of $S$, the potential must be
sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
iteration.

@@

In the rule {\sc L:Call}, accounts for the execution cost of both
$M_f$ for function calls and $M_r$ for \code{return} statements.  This
justifies that the {\sc L:Return} rule does not account for resource
consumption.  The pre- and postconditions $P_f$ and $Q_f$ are taken
from the function context $\Delta$.  The assertions in the context are
parametric with respect to both the function argument value and the
return value. This allows to specify a bound for a function whose
resource consumption depends on an input parameter.  The argument
parameter is instantiated by the call rule using the result of the
evaluation of the argument variables in the current state.
\iffull{Recall that we require that $P_f$ and $Q_f$ do not depend on
  the local variables on the call stack, that is, $\forall z \, \vec v
  \, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
  (\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.}{} To
transfer potential that depends on local variables of the callee from
the precondition $P$ to the postcondition $Q$, we use an assertion $A
: \Assn$ that is independent of global variables, that is, $\forall
\theta \, \gamma \, \gamma' . \, A (\theta,\gamma) =
A(\theta,\gamma')$.  It is still possible to express relations between
global and local variables using logical variables (see the
\iffull{following paragraph}{extended version} for details).

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.

>>>>

\iffull{In the rule {\sc L:Skip}, we do not have to account for any resource
consumption.  As a result, the precondition $Q$ can be any (potential)
function and we only have to make sure that we do not end up with more
potential.  Since the execution of \code{skip} leaves the program
state unchanged, we can simply use the precondition as postcondition.
The potential functions $B$ for the \code{break} and $R$ for the
\code{return} part of the postcondition are not reachable and can
therefore be arbitrary.}{}

@@

\iffull{The rule {\sc L:Update} is the standard assignment rule of Hoare
logic.  With the substitution $\state[x \mapsto \sem{e}_\state]$ in
the precondition we ensure that $Q$ evaluates to the same number as in
the postcondition.  We also require that have we the constant
potential $M_u + M_e(e)$ available in the precondition to pay for the
cost of the evaluation of $e$ and the update.

The rule {\sc L:Seq} rule is crucial to understand how the
quantitative Hoare logic works.  To account for early exits of
statements, we must ensure in the \code{break} part $B$ of $S_1$'s
judgement that the \code{break} part $B$ of of $S_1;S_2$ holds. The
same is true for the return part $R$ of the judgements for $S_1$ and
$S_2$.  The interaction between the actual pre- and postconditions is
analogous to standard Hoare logic.  In the postcondition of $S_1$ we
account for the cost $M_s$ of the execution of the sequence.

}
{
The rules {\sc L:Update} and {\sc L:Seq} correspond to the respective 
rules of standard Hoare logic.  
}
In the rule {\sc L:Loop}, the \code{break} part of the
loop body $S$ becomes the postcondition of the loop statement. We use
an arbitrary $B$ as the \code{break} part of the judgement for
$\code{loop}\, S$ since its operational semantics ensures that it can
only terminate with a \code{skip} or a \code{return}.  The
precondition $I$ of the loop is the loop invariant.  \iffull{That is why we
require to have potential $I$ available in the precondition of the
loop body $S$.}{}  In the postcondition of $S$, the potential must be
sufficient to pay for the invariant $I$ and the cost $M_l$ of the loop
iteration.

@@

In the rule {\sc L:Call}, accounts for the execution cost of both
$M_f$ for function calls and $M_r$ for \code{return} statements.  \iffull{This
justifies that the {\sc L:Return} rule does not account for resource
consumption.}{}  The pre- and postconditions $P_f$ and $Q_f$ are taken
from the function context $\Delta$.  The assertions in the context are
parametric with respect to both the function argument value and the
return value. This allows to specify a bound for a function whose
resource consumption depends on an input parameter.  The argument
parameter is instantiated by the call rule using the result of the
evaluation of the argument variables in the current state.
\iffull{Recall that we require that $P_f$ and $Q_f$ do not depend on
  the local variables on the call stack, that is, $\forall z \, \vec v
  \, \theta \, \theta' \, \gamma . \, P_f \, z \, \vec v \,
  (\theta,\gamma) = P_f \, z \, \vec v \, (\theta',\gamma)$.}{} To
transfer potential that depends on local variables of the callee from
the precondition $P$ to the postcondition $Q$, we use an assertion $A
: \Assn$ that is independent of global variables, that is, $\forall
\theta \, \gamma \, \gamma' . \, A (\theta,\gamma) =
A(\theta,\gamma')$.  It is still possible to express relations between
global and local variables using logical variables\iffull{ (see the
following paragraph for details).}{.}

Finally, we describe the rules which are not syntax directed.  There
are two weakening rules available in the quantitative Hoare logic.
The framing rule {\sc L:Frame} is designed to weaken a statement by
stating that if $S$ needs $P$ resources to run and leaves $Q$
resources available after its execution, then it can very well run
with $P + c$ resources and return $Q + c$ resources.  The consequence
{\sc L:Conseq} rule is directly imported from classical Hoare logic
except that instead of using the logical implication $\Rightarrow$ we
use the quantitative $\models$ that point-wise applies $\ge$.  \iffull{This
rule indeed weakens the statement since it requires more resource to
run the statement and yields less than what has been proved to be
available after its termination.}{}
